---
title: "Bactopia Analysis"
author: "Robert A Petit III and Tim Read"
date: "February 27, 2020"
output:
  html_document:
    df_print: paged
---
```{r}
library(dplyr)
library(ggplot2)
library(reshape2)
library(stringr)
library(tidyr)
library(tidyverse)
options(scipen=999)
DATA <- "../data"

# Read Inputs
sra_metadata <- read.delim(paste0(DATA,"/lactobacillus-results.txt")) %>% distinct(experiment_accession, .keep_all=TRUE)
sra_metadata$sample <- sra_metadata$experiment_accession
sequence_metrics = read.delim(paste0(DATA,"/summary/lactobacillus-report.txt"))
full_metrics = merge(sequence_metrics, sra_metadata, by='sample', all.x = TRUE)
full_metrics$year <- as.data.frame(select(full_metrics, first_public) %>% separate(first_public, into=c('year'), extra='drop'))$year
phyloflash <- read.delim(paste0(DATA,"/phyloflash/phyloflash-summary.txt"))
gtdb <- read.delim(paste0(DATA,"/gtdb/gtdbtk.summary.tsv"))
exclude <- read.delim(paste0(DATA,"/summary/lactobacillus-exclude.txt"))

#' write_plot
#'
#' A wrapper for to validate given vector is multiple ids and proper type. This
#' function should not be directly used by the user.
#'
#' @param plot_object A ggplot object
#' @param name Basename for the output PDF and PNG files
#' @param height The PDF height of the output (Default: 5)
#' @param width The PDF width of the object (Default: 12)
#'
#' @export
#' @return bool TRUE is multiple ids else FALSE.
write_plot <- function(plot_object, name, height = 5, width = 12) {
    pdf(paste0(name, ".pdf"), width=width, height=height)
    print(plot_object)
    dev_null <- dev.off()

    png(paste0(name, ".png"), width=width*100, height=height*100)
    print(plot_object)
    dev_null <- dev.off()
}
```

### Technologies
```{r}
table(sra_metadata$instrument_platform)
```

### By Year Plots
#### Submissions Per Year
```{r fig.width=12, fig.asp=0.4}
submissions <- select(full_metrics, year) %>% count(year)

p <- ggplot(data=submissions, aes(x=year, y=n)) +
    xlab("Year") +
    ylab("Count") +
    geom_bar(stat='identity') +
    geom_text(aes(label=n), vjust = -0.5) +
    scale_x_discrete(breaks = round(
        seq(min(submissions$year), max(submissions$year), by = 1),1
    )) +
    theme_bw() +
    theme(axis.text=element_text(size=12),
          axis.title=element_text(size=14,face="bold"))
p
```

# Sequence Summary
## Ranks
```{r}
table(sequence_metrics$rank)
```

### By Rank Table
#### Gold
```{r}
sequence_metrics %>%
    filter(rank == 'gold') %>%
    select(original_coverage, final_coverage, final_qual_mean, final_read_mean, total_contig) %>%
    map(summary)

summary(sequence_metrics[sequence_metrics$rank == 'gold',]$total_contig_length/sequence_metrics[sequence_metrics$rank == 'gold',]$estimated_genome_size)
```

#### Silver
```{r}
sequence_metrics %>%
    filter(rank == 'silver') %>%
    select(original_coverage, final_coverage, final_qual_mean, final_read_mean, total_contig) %>%
    map(summary)

summary(sequence_metrics[sequence_metrics$rank == 'silver',]$total_contig_length/sequence_metrics[sequence_metrics$rank == 'silver',]$estimated_genome_size)
```

#### Bronze
```{r}
sequence_metrics %>%
    filter(rank == 'bronze') %>%
    select(original_coverage, final_coverage, final_qual_mean, final_read_mean, total_contig) %>%
    map(summary)


summary(sequence_metrics[sequence_metrics$rank == 'silver',]$total_contig_length/sequence_metrics[sequence_metrics$rank == 'silver',]$estimated_genome_size)
```

#### Exclude
```{r}
sequence_metrics %>%
    filter(rank == 'exclude') %>%
    select(original_coverage, final_coverage, final_qual_mean, final_read_mean, total_contig) %>%
    map(summary)


summary(sequence_metrics[sequence_metrics$rank == 'silver',]$total_contig_length/sequence_metrics[sequence_metrics$rank == 'silver',]$estimated_genome_size)
```

# Taxonmic Comparisons
```{r}
sra_species <- select(full_metrics, experiment_accession, rank, estimated_genome_size, total_contig_length, scientific_name) %>%
                separate(scientific_name, into=c('genus', 'species'), extra='drop', fill='right') %>%
                unite('sra', c('genus', 'species'), sep=" ")
colnames(sra_species) <- c('sample', 'rank', 'estimated_genome_size', 'assembled_genome_size', 'sra')

phyloflash_species <- filter(phyloflash, !grepl("WARNING", message)) %>% 
                        select(sample, assembly_species) %>%
                        separate(assembly_species, into=c('genus', 'species'), extra='drop', fill='right') %>%
                        unite('phyloflash', c('genus', 'species'), sep=" ")

gtdb_species <- select(gtdb, user_genome, classification) %>%
                separate(classification, into=c('genus', 'species'), sep=";s__", extra='drop', fill='right') %>%
                select(user_genome, genus, species) %>%
                separate(genus, into=c('taxon', 'genus'), sep="g__", extra='drop', fill='right') %>%
                separate(species, into=c('temp_genus', 'species'), sep=" ", extra='drop', fill='right') %>%
                separate(temp_genus, into=c('temp_genus'), sep="_", extra='drop', fill='right') %>%
                separate(species, into=c('species'), sep="_", extra='drop', fill='right') %>%
                unite('species', c('temp_genus', 'species'), sep=" ") %>%
                select(user_genome, genus, species)


colnames(gtdb_species) <- c('sample', 'genus', 'species')
gtdb_species$gtdb <- ifelse(gtdb_species$species == ' NA', gtdb_species$genus,  gtdb_species$species)
gtdb_species$is_lacto = ifelse(grepl("Lactobacillus", gtdb_species$gtdb), TRUE, FALSE)
species <- merge(sra_species, 
    merge(gtdb_species, phyloflash_species, by=c("sample"), all.x = TRUE), 
    by=c("sample"), 
    all.x = TRUE
)
species$phyloflash[is.na(species$phyloflash)] <- 'unknown_phyloflash'
species$gtdb[is.na(species$gtdb)] <- 'unknown_gtdb'
species$all_agree <- species$sra == species$phyloflash & species$sra == species$gtdb
species$sra_phyloflash <- species$sra == species$phyloflash
species$sra_gtdb <- species$sra == species$gtdb
species$phyloflash_gtdb <- species$phyloflash == species$gtdb
species$is_lacto <- ifelse(grepl("Lactobacillus", species$gtdb), TRUE, FALSE)

```

### Samples with GTDB assignments
```{r}
nrow(gtdb_species)
```

### Non-Lactobacillus assignments by GTDB
```{r}
table(gtdb_species$is_lacto)
table(gtdb_species[!gtdb_species$is_lacto,]$gtdb)
```

### Mismatches between assignments
```{r}
# unknown_gtdb means gtdb wasn't run on the sample
table(species[species$gtdb != "unknown_gtdb",]$all_agree)
```

### Mismatches between 16S and GTDB
```{r}
# comapre samples with phyloflash and gtdb results
nrow(species %>% filter(gtdb != "unknown_gtdb" & phyloflash != "unknown_phyloflash"))
gtdb_phyloflash_mismatch <- species %>%
                              filter(gtdb != "unknown_gtdb" & phyloflash != "unknown_phyloflash" & !phyloflash_gtdb) %>%
                              select(gtdb, phyloflash) %>%
                              unite('mismatch', c('gtdb', 'phyloflash'), sep=";")
nrow(gtdb_phyloflash_mismatch)
table(gtdb_phyloflash_mismatch)
```

```{r}
estimated_sizes <- select(species, gtdb, estimated_genome_size) %>%
                    filter(gtdb != "unknown_gtdb") %>%
                    group_by(gtdb) %>%
                    summarise_all(funs(length, min, median, mean, max, max-min, sd))
colnames(estimated_sizes) <- c('species', 'count', 'min', 'median', 'mean', 'max', 'range', 'standard_deviation')
filter(estimated_sizes, count >50) %>% arrange(desc(count))
```

```{r}
assembly_sizes <- select(species, gtdb, assembled_genome_size) %>%
                    filter(gtdb != "unknown_gtdb") %>%
                    group_by(gtdb) %>%
                    summarise_all(funs(length, min, median, mean, max, max-min, sd))
colnames(assembly_sizes) <- c('species', 'count', 'min', 'median', 'mean', 'max', 'range', 'standard_deviation')
filter(assembly_sizes, count >50) %>% arrange(desc(count))
```
# Supplementary Tables and Figures
### Supplementary Table 1 - Supplementary Table 1 - Lactobacillus excluded from analysis
```{r}
exclude %>% select(sample, reason)
```

### Supplementary Table 2 - Samples with a non-Lactobacillus taxonomic classification
```{r}
species %>% filter(gtdb != 'unknown_gtdb' & !is_lacto ) %>% select(sample, sra, gtdb) 
```

### Supplementary Figure 2 - Sequencing quality ranks per year 2011–2019 of Lactobacillus genome projects. Submission Ranks
```{r Supplementary Figure 2, fig.width=12, fig.asp=0.5}
melted <- select(full_metrics, rank, year) %>% 
            filter(rank != 'exclude') %>% 
            mutate(rank = sub("(.)", "\\U\\1", rank, perl=TRUE)) %>% 
            count(year, rank)
p <- ggplot(data=melted, aes(x=year, y=n, fill=rank, group=rank, label=year)) +
    xlab("Year") +
    ylab("Count") +
    geom_bar(stat='identity', position='dodge') +
    geom_text(aes(label=n), vjust = -0.5, position = position_dodge(.9)) +
    scale_fill_manual(values=c("#CD7F32", "#D4AF37", "#C0C0C0" )) +
    scale_x_discrete(breaks = round(
        seq(min(melted$year), max(melted$year), by = 1), 1
    )) +
    theme_bw() +
    theme(axis.text=element_text(size=12),
          axis.title=element_text(size=14,face="bold"),
          legend.title = element_blank())
    
p

write_plot(p, paste0(getwd(), '/../preprint/supplementary-quality-by-year'))
```


### Supplementary Figure 3 - Comparison of estimated genome size and assembled genome size.
```{r Supplementary Figure 3, fig.width=12, fig.asp=0.5}
p <- ggplot(full_metrics, aes(x = estimated_genome_size, y = total_contig_length, color=rank)) +
    xlab("Estimated Genome Size") +
    ylab("Assembled Genome Size") +
    geom_point() +
    geom_abline(intercept = 0, slope = 1) + 
    scale_color_manual(values = c("#CD7F32", "#000000", "#D4AF37", "#C0C0C0" )) +
    theme_bw() +
    theme(axis.text=element_text(size=12),
          axis.title=element_text(size=14,face="bold"),
          legend.title = element_blank())
    
p

write_plot(p, paste0(getwd(), '/../preprint/figures/supplementary-genome-size-assembly-vs-estimate'))
```


### Supplementary Figure 4 - Assembled genome sizes are consistent within species.
```{r Supplementary Figure 4, fig.width=12, fig.asp=0.5}
species_genomes <- select(species, gtdb, assembled_genome_size) %>% 
    filter(gtdb != "unknown_gtdb" & !is.na(gtdb)) %>%
    group_by(.data = ., gtdb) %>%
    mutate(.data = ., n = n()) %>%
    separate(gtdb, c("genus", "species"), sep=" ", extra='drop', fill='right') %>%
    filter(n>50)

species_genomes$label <- paste0("L. ", species_genomes$species, " (n=", species_genomes$n, ")")

p <- ggplot(species_genomes, aes(x = label, y = assembled_genome_size)) +
    xlab("Species") +
    ylab("Assembled Size (bp)") +
    geom_boxplot() +
    theme_bw() +
    theme(axis.text=element_text(size=10),
          axis.text.x = element_text(face = 'italic'),
          axis.title=element_text(size=14,face="bold"),
          legend.title = element_blank()) 
    
p

write_plot(p, paste0(getwd(), '/../preprint/supplementary-consistent-genome-size'))
```

### Antibiotic Resistance

```{r}
AMR_finder <- read.delim(paste0(DATA,"/summary/amrfinder/amrfinder-gene-summary.txt"))  %>%
  rename(sample = sample_name) %>%
  mutate_at(2:11,as.logical)

CARD <- read.delim(paste0(DATA,"/summary/ariba/ariba-card-summary.txt")) %>%
  rename(sample = sample_name) %>%
  mutate_at(2:48,as.logical)
```

```{r}
Lacto_AMR <- inner_join(gtdb_species,AMR_finder,by="sample") %>%
  filter(is_lacto == TRUE) %>%
  select(-c(is_lacto)) %>%
  mutate(ANY = ifelse(rowSums(.[5:14]),TRUE,FALSE))

Lacto_CARD <- inner_join(gtdb_species,CARD,by="sample") %>%
  filter(is_lacto == TRUE) %>%
  select(-c(is_lacto)) %>%
  mutate(ANY = ifelse(rowSums(.[5:51]),TRUE,FALSE))
```


```{r}
count_AMR <- Lacto_AMR %>%
  group_by(gtdb) %>% 
  summarise(count = n())

Lacto_AMR_summary <- Lacto_AMR %>% 
  group_by(gtdb) %>% 
  summarise_if(is.logical,sum,na.rm = T) 

Lacto_AMR_summary2 <- inner_join(count_AMR,Lacto_AMR_summary, by = "gtdb")
colSums(Lacto_AMR_summary2[,2:13])
```

```{r}
Lacto_AMR_summary2$total <- rowSums(Lacto_AMR_summary2[3:12])
filter(Lacto_AMR_summary2, total >5) %>% arrange(desc(total))
```


```{r}
# Ariba was not run on the single-end sampels, hence different counts then
# the AMRFinder+ counts
count_CARD <- Lacto_CARD %>%
  group_by(gtdb) %>% 
  summarise(count = n())

CARD_summary <- Lacto_CARD %>% 
  group_by(gtdb) %>% 
  summarise_if(is.logical,sum,na.rm = T) 

CARD_summary2 <- inner_join(count_CARD,CARD_summary, by = "gtdb")
CARD_numbers <- colSums(CARD_summary2[,2:50]) 
CARD_numbers[CARD_numbers > 0]
```

# Session Info
```{r}
sessionInfo()
```